---
title: "CODING MUTATIONS"
author: "ElsaB"
date: "18/05/29"
output: html_document
---

```{r, echo=FALSE, results="hide", include=FALSE}
source("../../../src/utils/ggstyles.R")
```

# Coding mutations

Here we will only select the coding mutations (including splicing) from the IMPACT dataset.

```{r echo=FALSE}
dd = read.table("../../../data/all_IMPACT_mutations_180508.txt", sep="\t", stringsAsFactors=F, header=T)
```

```{r echo=TRUE}
# RAW DATA:
# size
nrow(dd)
# variant effect versus assignment:
table(dd$Consequence, dd$confidence_class)
```

```{r echo=TRUE}
# FILTERING:
# coding mutations
keep.coding = c("frameshift_deletion","frameshift_insertion",
	      "nonframeshift_deletion","nonframeshift_insertion",
	      "nonsynonymous_SNV",
	      "splicing","splicing_noncanonical",
	      "stopgain_SNV","stoploss_SNV",
	      "Translation_Start_Site"
	      )
hh = dd[which(dd$Consequence %in% keep.coding),]
```

```{r echo=TRUE}
# size after type filtering:
nrow(hh)
# assignement:
table(hh$Consequence, hh$confidence_class)
# del ins snv:
table(hh$Variant_Type)
```

```{r echo=FALSE}
hh$variant.key = paste(hh$Chromosome,
		       hh$Start_Position,
		       hh$Reference_Allele,
		       hh$Tumor_Seq_Allele2,
		       sep="_"
		       )
```

# Contamination

Remove contaminated samples:

```{r echo=TRUE}
# cutoff
hh = hh[hh$minor_contamination <= 0.01,]
# size
nrow(hh)
```


# Coding curated mutations

Here we select the coding mutations that have been curated as being somatic / non-somatic.

Most of the coding mutations have been curated, only a few splicing/stoploss are in the unknowns.

```{r echo=TRUE}
# keep the curated mutations
hh = hh[hh$confidence_class != "UNKNOWN",]
# somatic versus non-somatic label
hh$label = "somatic"
hh$label[hh$confidence_class=="UNLIKELY" | hh$confidence_class=="OK_NOT_SO"] = "non-somatic"
```

The dataset is highly unbalanced:

```{r echo=TRUE}
table(hh$label)
```

```{r echo=TRUE}
# consequence versus somatic
table(hh$Consequence, hh$label)
# variant type versus somatic:
table(hh$Variant_Type, hh$label)
```

## Frequency in normals

We look at the frequency of somatic / non-somatic variants in the pool of normals.

```{r echo=FALSE}
hh$occurence_in_normals[which(hh$occurence_in_normals=="0")] = "0;0"
hh$occurence_in_normals[which(hh$occurence_in_normals=="")] = "0;0"
hh$freq_normals = zaza=sapply(hh$occurence_in_normals, function(x) 100*as.numeric(strsplit(x,split=";")[[1]][2]))
hh$in_normals = "no"
hh$in_normals[hh$freq_normals!=0] = "yes"
```

```{r, echo=FALSE}
mytwolabel = c("#f1a340","#998ec3")
# density
ggplot(hh, aes(x=freq_normals, fill=label)) + geom_density(alpha=.9) + theme1 + topleg + nolegtitle + scale_x_sqrt() + scale_y_sqrt() + scale_fill_manual(values=mytwolabel)
# barplot
ggplot(hh, aes(x=in_normals, fill=label)) + geom_bar(alpha=.9) + theme1 + topleg + nolegtitle + scale_fill_manual(values=mytwolabel)
```

```{r echo=TRUE}
iwhy = which(hh$label=="somatic" & hh$in_normals=="yes")
summary(hh[iwhy,"freq_normals"])
length(iwhy)
```

**what happenned with those supposely somatic calls that are present in the normals as well???**


## Tumor VAF and DIRPROP

We investigate VAF and strand bias.

First, reproduce the VAF as a sanity check:

```{r echo=TRUE}
ddvaf = data.frame(raw_t_vaf = hh$t_vaf,
		   do_t_vaf = as.numeric(hh$t_alt_count / hh$t_depth),
		   go_t_vaf = as.numeric((hh$t_alt_neg_count+hh$t_alt_plus_count) / hh$t_depth)
		   )
# sanity check #1:
ggplot(ddvaf) + geom_point(aes(x=raw_t_vaf, y=do_t_vaf)) + theme1
# sanity check #2:
ggplot(ddvaf) + geom_point(aes(x=raw_t_vaf, y=go_t_vaf)) + theme1
```

**There is a couple of outliers where we are not able to reproduce the VAF based on the raw counts.**
This might be related to different mapping quality thresholds...


Distribution of VAF for the somatic / non-somatic categories:


```{r echo=FALSE, results="hide"}
# density of the tumor VAF
ggplot(hh, aes(x=t_vaf, fill=label)) + geom_density(alpha=.7) + theme1 + topleg + nolegtitle + scale_fill_manual(values=mytwolabel)
```

As expected, the non-somatic density has a peak at 0.5 (ie the likely germline calls),
and a peak at low VAF (ie the likely artefacts calls).

Second, we look at the strand bias:

```{r echo=TRUE}
hh$DIRPROP = pmin(hh$t_alt_neg_count,hh$t_alt_plus_count) / pmax(hh$t_alt_neg_count,hh$t_alt_plus_count)
```

Distribution of strand ratio for the somatic / non-somatic categories:


```{r echo=FALSE, results="hide"}
# density of the tumor DIRPROP
ggplot(hh, aes(x=DIRPROP, fill=label)) + geom_density(alpha=.7) + theme1 + topleg + nolegtitle + scale_fill_manual(values=mytwolabel)
```

As expected, the non-somatic density has a heavier low strand ratio tail (likely from the artefact calls),
although this is not very pronounced.
